Fundamental measure theory for hard polyhedra.

### 2013-02-12 papers/polyhedra/paper.tex

\subsection{Ideal gas free energy}
To address this, we will
first consider the free energy of an ideal gas.
\begin{widetext}
\begin{align}
  A_{\textit{ideal}} &=
  k_BT \int
  \left(n(\rr,\rot)\ln(\Lambda^3\Theta^3 n(\rr,\rot)) - n(\rr,\rot)
  \right) d\rr d\rot \label{eq:Aideal}
  \\
  &=
  k_BT \sum_{lmm'} \int
  n_{lmm'}(\rr)D_{lmm'}(\rot)\left(
  \ln\left(\Lambda^3\Theta^3 \sum_{l_1m_1m'_1} n_{l_1m_1m'_1}(\rr)D_{l_1m_1m'_1}(\rot)\right) - 1
  \right) d\rr d\rot
  \\
  &=
  k_BT \int n_{000}(\rr)
  \left(\ln\left(\Lambda^3\Theta^3 \sum_{lmm'} n_{lmm'}(\rr)D_{lmm'}(\rot)\right) - 1
  \right) d\rr d\rot
  \\&+
  k_BT \sum_{l>1,mm'} \int
  n_{lmm'}(\rr)D_{lmm'}(\rot)
  \ln\left(\Lambda^3\Theta^3 \sum_{l_1m_1m'_1} n_{l_1m_1m'_1}(\rr)D_{l_1m_1m'_1}(\rot)\right)
  d\rr d\rot
  \\
  &= A_{000} + \sum_{l>1,mm'} A_{lmm'} - k_BT\int n_{000}(\rr)d\rr
\end{align}
Here I have broken the $000$ term out of the outer summation over
$lmm'$, since this one is special, since it is independent of $\rot$.
We'll tackle that integral first, since it is a tad simpler.
\begin{align}
  A_{000} &= k_BT\int
  n_{000}(\rr)\ln\left(
  \sum_{lmm'} n_{lmm'}(\rr)D_{lmm'}(\rot)
  \right)
  d\rr d\rot
\end{align}
We can simplify this integral a bit if we write the log as a power
series...
\begin{align}
  A_{000}
  &= k_BT\int
  n_{000}(\rr)\ln\left(
  n_{000}(\rr)\left(1 + \frac{\sum_{lmm'} n_{lmm'}(\rr)D_{lmm'}(\rot)}{n_{000}(\rr)}\right)
  \right)
  d\rr d\rot
  \\
  &= k_BT\int
  n_{000}(\rr)\ln n_{000}(\rr)
  \\&+k_BT\int
  n_{000}(\rr)\ln\left(1 + \frac{\sum_{lmm'} n_{lmm'}(\rr)D_{lmm'}(\rot)}{n_{000}(\rr)}\right)
  d\rr d\rot
  \\
  &= k_BT\int
  n_{000}(\rr)\ln n_{000}(\rr)
  \\&+k_BT\int
  n_{000}(\rr)\left(\frac{\sum_{lmm'} n_{lmm'}(\rr)D_{lmm'}(\rot)}{n_{000}(\rr)}\right)
  d\rr d\rot
  \\&-\frac{k_BT}{2!}\int
  n_{000}(\rr)\left(\frac{\sum_{lmm'} n_{lmm'}(\rr)D_{lmm'}(\rot)}{n_{000}(\rr)}\right)^2
  d\rr d\rot
  \\&+\cdots
\end{align}
Now the linear term in the power series will vanish, since the
summation is over $l>0$ only.  In the quadratic term, only a few terms
will survive, and we'll end up with
\begin{align}
  A_{000}
  &=
  k_BT\int
  n_{000}(\rr)\ln n_{000}(\rr)d\rr
  \\&\quad-\frac{k_BT}{2!}\int
  \frac{\sum_{lmm'} \frac{2l+1}{8\pi^2} n_{lmm'}(\rr)n_{l,-m,-m'}}{n_{000}(\rr)}
  d\rr
  \\&\quad+\cdots
  \\
  &=
  k_BT\int
  n_{000}(\rr)\ln n_{000}(\rr)d\rr
  \\&\quad-\frac{k_BT}{2!}\int
  \frac{\sum_{lmm'} \frac{2l+1}{8\pi^2} |n_{lmm'}(\rr)|^2}{n_{000}(\rr)}
  d\rr
  \\&\quad+\cdots
\end{align}
Now an interesting question is whether we can mimic this power series
with a log, such as
\begin{align}
  A_{000}
  &=
  k_BT\int
  n_{000}(\rr)\ln \left( \sqrt{n_{000}(\rr)^2 + \sum_{lmm'}
    c_ln_{lmm'}(\rr)^2} \right)d\rr
  \\
  &= 
  k_BT\int
  n_{000}(\rr)\ln n_{000}(\rr)d\rr
  \\&\quad+
  k_BT\int n_{000}(\rr)\ln\left( \sqrt{1 + \frac{\sum_{lmm'}
    c_ln_{lmm'}(\rr)^2}{n_{000}(\rr)^2}} \right)d\rr
  \\
  &=
  k_BT\int
  n_{000}(\rr)\ln n_{000}(\rr)d\rr
  \\&\quad+
  k_BT\int \frac12 \frac{\sum_{lmm'}
    c_ln_{lmm'}(\rr)^2}{n_{000}(\rr)} d\rr
  \\&\quad+\cdots
\end{align}
from which we can conclude that to get an answer correct to the
quadratic order, we need
\begin{align}
  c_l &= -\frac{2l+1}{8\pi^2}
\end{align}
which seems all right except for the sign:  I expect that the free
energy should increase as we increase $n_{lmm'}$, since it means the
system is becoming more ordered, and it looks like to this quadratic
term, the free energy decreases... except that we're only looking at
$A_{000}$ right now, and thus are neglecting a possibly significant
increase due to the other terms...  so perhaps we should call this all
right, and move on for now.

The $A_{lmm'}$ terms are a bit different, but not too different:
\begin{align}
  A_{lmm'} &=
  k_BT \int n_{lmm'}(\rr) D_{lmm'}(\rot)
  \ln\left(\sum_{l_1m_1m'_1} n_{l_1m_1m'_1}(\rr)
  D_{l_1m_1m'_1}(\rot)\right)
  d\rr d\rot
  \\
  &=
  k_BT \int n_{lmm'}(\rr) D_{lmm'}(\rot)
  \ln\left(1 + \frac{\sum_{l_1m_1m'_1} n_{l_1m_1m'_1}(\rr)
  D_{l_1m_1m'_1}(\rot)}{n_{000}(\rr)}\right)
  d\rr d\rot
  \\
  &=
  k_BT \int n_{lmm'}(\rr) D_{lmm'}(\rot)
  \left(\frac{\sum_{l_1m_1m'_1} n_{l_1m_1m'_1}(\rr)
  D_{l_1m_1m'_1}(\rot)}{n_{000}(\rr)}+\cdots\right)
  d\rr d\rot
  \\
  &=
  k_BT \int n_{lmm'}(\rr) \frac{8\pi^2}{2l+1} \frac{n_{l,-m,-m'}(\rr)}{n_{000}(\rr)}
  d\rr + \cdots
  \\
  &=
  k_BT \int \frac{8\pi^2}{2l+1} \frac{|n_{lmm'}(\rr)|^2}{n_{000}(\rr)}
  d\rr + \cdots
\end{align}
It is interesting how similar the $A_{lmm'}$ terms are to the
quadratic terms in $A_{000}$, off by a factor of $-\frac12$.  This
leads me to wonder about the possibility of writing
\begin{align}
  A_{\textit{ideal}} &\approx
  k_BT \int n_{000}(\rr)\left( \ln\left(
  \sqrt{\sum_{lmm'}\frac{2l+1}{16\pi^2}|n_{lmm'}|^2}
  \right) - 1\right)d\rr
\end{align}
This would be correct up to the quadratic terms, and doesn't require
any rotational integrals.  It also seems liable to be reasonable up to
higher order terms, although I have no guarantee of that.  An
interesting excersize would be to take this out to the fourth order,
and also take the exact expression out to the fourth order and see how
they compare.  They will actually differ in the third order terms, I
expect, since the exact expression will indeed have nonzero third
order terms.

So we have two plausible options that I can see: we could either go
with this logarithmic approximation, or we could just go out to second
order.  The latter is probably more defensible in terms of
perturbation theory, and would give us something like
\begin{align}
  A_{\textit{ideal}} &\approx
  k_BT \int \left(n(\rr)\ln n(\rr) - n(\rr) +
  \sum_{lmm'}\frac{2l+1}{16\pi^2}\frac{|n_{lmm'}|^2}{n_{000}(\rr)}
  \right)d\rr
\end{align}
where I have substituted the notation $n(\rr)$ for $n_{000}(\rr)$ to
save space.

\end{widetext}
\subsection{Finding the orientational moments}
We find the orientational moments $n_{lmm'}(\rr)$ by minimizing the
free energy.  Let's begin by optimizing the free energy with respect
to nonzero angular momenta:
\begin{align}
  0&=\frac{\delta A}{\delta n_{lmm'}(\rr)} \\
  &= \frac{2l+1}{8\pi^2}n_{lmm'}(\rr)
  + k_BT \int\sum_\alpha
  \frac{\partial \Phi(\rr')}{\partial n_\alpha(\rr')}
  \frac{\delta n_\alpha(\rr')}{\delta n_{lmm'}(\rr)} d\rr'
  \\
  &=
   \frac{2l+1}{8\pi^2}n_{lmm'}(\rr)
  \\&\quad+ k_BT \int\sum_\alpha
  \frac{\partial \Phi(\rr')}{\partial n_\alpha(\rr')}
  \tilde{w}_\alpha^{lm}(|\rr-\rr'|)Y_{lm'}(\widehat{\rr-\rr'}) d\rr'
\end{align}
Solving for $n_{lmm'}(\rr)$, we find that
\begin{align}\label{eq:nlmm'}
  n_{lmm'}(\rr) &= \frac{8\pi^2k_BT}{2l+1}
  \int\sum_\alpha
  \frac{\partial \Phi(\rr')}{\partial n_\alpha(\rr')}
  \tilde{w}_\alpha^{lm}(|\rr-\rr'|)Y_{lm'}(\widehat{\rr-\rr'}) d\rr'
\end{align}
I note that this equation for $n_{lmm'}$ is approximate, since we made
an approximation when we neglected higher order terms in the expansion
of the log in the ideal gas free energy.  Thus in cases where there is
relatively little orientational ordering, it ought to be an accurate
equation.

In cases where $n_{lmm'} \approx n_{000}$, we could reasonbly become
concerned that $n(\rr, \rot)$ could become negative for some
orientations, which would be unphysical.  If this becomes a problem,
we would want to address it by including nonlinear effects of the
$\ln$ in Equation~\ref{eq:Aideal}.  For now, let's stick with the
linear approximation, and hope for the best.  If we find that
$n_{lmm'} > n_{000}$ in practice, we would want to revisit this decision.

>>>>

\subsection{Ideal gas free energy}
To address this, we will
first consider the free energy of an ideal gas.
\begin{widetext}
\begin{align}
  A_{\textit{ideal}} &=
  k_BT \int
  \left(n(\rr,\rot)\ln(\Lambda^3\Theta^3 n(\rr,\rot)) - n(\rr,\rot)
  \right) d\rr d\rot \label{eq:Aideal}
  \\
  &=
  k_BT \sum_{lmk} \int
  n_{lmk}(\rr)D_{lmk}(\rot)\left(
  \ln\left(\Lambda^3\Theta^3 \sum_{l'm'k'} n_{l'm'k'}(\rr)D_{l'm'k'}(\rot)\right) - 1
  \right) d\rr d\rot
  \\
  &=
  k_BT 8\pi^2 n_{000}(\rr)( \ln(\Lambda^3\Theta^3) - 1) +
  k_BT \sum_{lmk} \int
  n_{lmk}(\rr)D_{lmk}(\rot)
  \ln\left(n_{000}(\rr) + \sum_{\substack{l'>0\\m'k'}} n_{l'm'k'}(\rr)D_{l'm'k'}(\rot)\right)
  d\rr d\rot
  \\
  &=
  k_BT 8\pi^2 n_{000}(\rr)( \ln(\Lambda^3\Theta^3n_{000}(\rr)) - 1) +
  k_BT \sum_{lmk} \int
  n_{lmk}(\rr)D_{lmk}(\rot)
  \ln\left(1 + \sum_{\substack{l'>0\\m'k'}} \frac{n_{l'm'k'}}{n_{000}(\rr)}(\rr)D_{l'm'k'}(\rot)\right)
  d\rr d\rot
  \\
  &=
  k_BT 8\pi^2 n_{000}(\rr)( \ln(\Lambda^3\Theta^3n_{000}(\rr)) - 1)
  \\&\quad+
  k_BT \sum_{lmk} \int
  n_{lmk}(\rr)D_{lmk}(\rot)
  \left(\sum_{\substack{l'>0\\m'k'}}
  \frac{n_{l'm'k'}(\rr)}{n_{000}(\rr)}D_{l'm'k'}(\rot)
  - \frac12 \left(\sum_{\substack{l'>0\\m'k'}}
    \frac{n_{l'm'k'}(\rr)}{n_{000}(\rr)}D_{l'm'k'}(\rot)
  \right)^2
  + \cdots \right)
  d\rr d\rot
  \\
  &=
  k_BT 8\pi^2 n_{000}(\rr)( \ln(\Lambda^3\Theta^3n_{000}(\rr)) - 1)
  %\\&\quad
  +
  k_BT \sum_{l>0mk}
  \frac{8\pi^2}{2l+1}\frac{|n_{l'm'k'}(\rr)|^2}{n_{000}(\rr)}
  - \frac12 \frac{8\pi^2}{2l+1}\frac{|n_{l'm'k'}(\rr)|^2}{n_{000}(\rr)}
  + \cdots \label{eq:integrated}
  \\
  &=
  k_BT 8\pi^2 n_{000}(\rr)( \ln(\Lambda^3\Theta^3n_{000}(\rr)) - 1)
  +
  k_BT \sum_{l>0mk}
  \frac{4\pi^2}{2l+1}\frac{|n_{l'm'k'}(\rr)|^2}{n_{000}(\rr)}
  + \cdots \\
  &\approx
  k_BT 8\pi^2 n_{000}(\rr)( \ln(\Lambda^3\Theta^3n_{000}(\rr)) - 1)
  -
  \frac{n_{000}(\rr)}{2}k_BT \ln\left(1 - \sum_{l>0mk}
  \frac{8\pi^2}{2l+1}\frac{|n_{l'm'k'}(\rr)|^2}{n_{000}(\rr)}
  \right) \label{eq:approxideal}
\end{align}
\end{widetext}
In Equation~\ref{eq:integrated} we have used the fact that
$D_{000}(\rot)=1$, and because $n(\rr,\rot)$ is real, $n_{l,-m,-k} =
n_{lmk}^*$.  In Equation~\ref{eq:approxideal}, we have constructed an
approxmation to the ideal gas free energy which has the correct
first-order response in the orientational order $n_{l>0,mk}(\rr)$ and
diverges (as it should) when $n_{l>0,mk}(\rr) \gg n_{000}(\rr)$, which
would lead to the nonsense result that $n(\rr,\rot) < 0$.  We can tune
the divergence a bit by moving constant factors (e.g. $8\pi^2$ into
our out of the logarithm.

\subsection{Finding the orientational moments}
We find the orientational moments $n_{lmm'}(\rr)$ by minimizing the
free energy.  Let's begin by optimizing the free energy with respect
to nonzero angular momenta:
\begin{align}
  0&=\frac{\delta A}{\delta n_{lmm'}(\rr)} \\
  &= \frac{2l+1}{8\pi^2}n_{lmm'}(\rr)
  + k_BT \int\sum_\alpha
  \frac{\partial \Phi(\rr')}{\partial n_\alpha(\rr')}
  \frac{\delta n_\alpha(\rr')}{\delta n_{lmm'}(\rr)} d\rr'
  \\
  &=
   \frac{2l+1}{8\pi^2}n_{lmm'}(\rr)
  \\&\quad+ k_BT \int\sum_\alpha
  \frac{\partial \Phi(\rr')}{\partial n_\alpha(\rr')}
  \tilde{w}_\alpha^{lm}(|\rr-\rr'|)Y_{lm'}(\widehat{\rr-\rr'}) d\rr'
\end{align}
Solving for $n_{lmm'}(\rr)$, we find that
\begin{align}\label{eq:nlmm'}
  n_{lmm'}(\rr) &= \frac{8\pi^2k_BT}{2l+1}
  \int\sum_\alpha
  \frac{\partial \Phi(\rr')}{\partial n_\alpha(\rr')}
  \tilde{w}_\alpha^{lm}(|\rr-\rr'|)Y_{lm'}(\widehat{\rr-\rr'}) d\rr'
\end{align}
I note that this equation for $n_{lmm'}$ is approximate, since we made
an approximation when we neglected higher order terms in the expansion
of the log in the ideal gas free energy.  Thus in cases where there is
relatively little orientational ordering, it ought to be an accurate
equation.

In cases where $n_{lmm'} \approx n_{000}$, we could reasonbly become
concerned that $n(\rr, \rot)$ could become negative for some
orientations, which would be unphysical.  If this becomes a problem,
we would want to address it by including nonlinear effects of the
$\ln$ in Equation~\ref{eq:Aideal}.  For now, let's stick with the
linear approximation, and hope for the best.  If we find that
$n_{lmm'} > n_{000}$ in practice, we would want to revisit this
decision, and make use of something like
Equation~\ref{eq:approxideal}.
